---
title: "Variables over Time"
toc: true
execute:
  echo: false
  warning: false
  message: false
---

{{< include _setup.qmd >}}

### Number of Fire Disasters per Region over Time

```{r events_over_time, fig.height = 8, fig.width = 8}
plotme(wfbz_crit, 'Events', rel = FALSE) +
  plot_spacer() +
  plotme(wfbz_crit, 'Events', rel = TRUE) +
  plot_layout(ncol = 1, heights = c(5, 1, 5), guides = 'collect')
ggsave(
  here('figures/events_over_time.pdf'),
  last_plot(),
  width = 8,
  height = 8
)
```


### Number of Fatalities per Region over Time

The five deadliest fires in our data set are labeled.


```{r fatalities_over_time, fig.height = 8, fig.width = 8}
plotme(wfbz_crit, 'Fatalities', rel = FALSE) +
  annotate(
    "text",
    x = 2023,
    y = 50,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Lahaina Fire (HI) ———"
  ) +
  annotate(
    "text",
    x = 2018,
    y = 80,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Camp Fire (CA) ———"
  ) +
  annotate(
    "text",
    x = 2017,
    y = 50,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Tubbs Fire (CA) ———"
  ) +
  annotate(
    "text",
    x = 2013,
    y = 19,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Yarnell Hill Fire (AZ) ———"
  ) +
  annotate(
    "text",
    x = 2020,
    y = 20,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "North Complex Fire (CA) ———"
  ) +
  plot_spacer() +
  plotme(wfbz_crit, 'Fatalities', rel = TRUE) +
  plot_layout(ncol = 1, heights = c(5, 1, 5))
ggsave(
  here('figures/fatalities_over_time.pdf'),
  last_plot(),
  width = 8,
  height = 8
)
```


### Number of Structures Destroyed per Region over Time

The five fires in our data set with the most destroyed structures are labeled.

```{r structures_over_time, fig.height = 8, fig.width = 16}
plotme(wfbz_crit, 'Structures Destroyed', rel = FALSE) +
  annotate(
    "text",
    x = 2018,
    y = 154 * 100,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Camp Fire (CA) ———"
  ) +
  annotate(
    "text",
    x = 2017,
    y = 85 * 100,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Tubbs Fire (CA) ———"
  ) +
  annotate(
    "text",
    x = 2023,
    y = 22 * 100,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Lahaina Fire (HI) ———"
  ) +
  annotate(
    "text",
    x = 2015,
    y = 25 * 100,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Valley Fire (CA) ———"
  ) +
  annotate(
    "text",
    x = 2018,
    y = 140 * 100,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Woolsey (CA) ———"
  ) +
  plot_spacer() +
  plotme(wfbz_crit, 'Structures Destroyed', rel = TRUE) +
  plot_layout(ncol = 1, heights = c(5, 1, 5))
ggsave(
  here('figures/structures_over_time.pdf'),
  last_plot(),
  width = 16,
  height = 8
)
```


### Number of FEMA Declarations per Region over Time

```{r fema_over_time, fig.height = 8, fig.width = 16}
plotme(wfbz_crit, 'FEMA Declarations (n)', rel = FALSE) +
  plot_spacer() +
  plotme(wfbz_crit, 'FEMA Declarations (n)', rel = TRUE) +
  plot_layout(ncol = 1, heights = c(5, 1, 5))
ggsave(
  here('figures/fema_over_time.pdf'),
  last_plot(),
  width = 16,
  height = 8
)
```

### Burned Area over Time

_i.e. the total area of all wildfire footprints per region/year (even if a fire affected the same spot twice in the year)_

```{r burned_area_over_time, fig.height = 8, fig.width = 16}
plotme(wfbz_crit, 'Burned Area', rel = FALSE, unit = 'km²') +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0)),
    labels = scales::comma
  ) +
  plot_spacer() +
  plotme(wfbz_crit, 'Burned Area', rel = TRUE) +
  plot_layout(ncol = 1, heights = c(5, 1, 5))
ggsave(
  here('figures/burned_area_over_time.pdf'),
  last_plot(),
  width = 16,
  height = 8
)
```

### Average Fire Size over Time

_i.e. the mean area of the wildfire footprints in a region/year_


```{r fire_size_over_time, fig.height = 4, fig.width = 4}
plotme(wfbz_crit, 'Mean Fire Size', rel = FALSE, unit = 'km²') +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0)),
    labels = scales::comma
  ) +
  plot_layout(ncol = 1)
ggsave(
  here('figures/fire_size_over_time.pdf'),
  last_plot(),
  width = 4,
  height = 4
)
```

## Share of WUI Fires over Time

```{r wui_over_time, fig.height = 12, fig.width = 14}
plotme(wfbz_crit, 'WUI', stack = FALSE) +
  xlab('')
ggsave(
  here('figures/wui_over_time.pdf'),
  last_plot(),
  width = 14,
  height = 12
)
```
