---
title: "Variables over Time"
toc: true
execute:
  echo: false
  warning: false
  message: false
---

{{< include _setup.qmd >}}

```{r map_setup}
hexgrid <- st_make_grid(states_pop, square = FALSE, n = c(500, 500)) %>%
  st_as_sf() %>%
  mutate(hexid = row_number()) %>%
  st_filter(states_pop)


wfbz_hex <- hexgrid %>%
  st_join(
    wfbz %>%
      mutate(
        usfs_region = wfbz_region,
        usfs_region_area = wfbz_region_area,
        usfs_region_pop = wfbz_region_pop
      ),
    left = FALSE
  ) %>%
  filter(wildfire_community_intersect) %>%
  group_by(hexid, geometry = x) %>%
  summarize(
    `Events` = n(),
    `Fatalities` = sum(wildfire_max_civil_fatalities, na.rm = TRUE),
    `Fatalities per 100K` = 100000 *
      sum(wildfire_max_civil_fatalities / usfs_region_pop, na.rm = TRUE),
    `Structures Destroyed` = sum(wildfire_struct_destroyed, na.rm = TRUE),
    `FEMA Declarations (n)` = sum(wildfire_fema_dec, na.rm = TRUE),
    `Burned Area` = sum(wildfire_area, na.rm = TRUE),
    `Burned Area Percent` = 100 *
      sum(wildfire_area / usfs_region_area, na.rm = TRUE),
    `WUI` = sum(str_detect(wildfire_wui, '(intermix|interface)'), na.rm = TRUE),
    `WUI Percent` = 100 *
      sum(str_detect(wildfire_wui, '(interface|intermix)'), na.rm = TRUE) /
      n(),
    `Mean Fire Size` = mean(wildfire_area, na.rm = TRUE),
    .groups = 'drop'
  )

mapme <- function(
  dat,
  var,
  flab = \(.x) scales::comma(.x, accuracy = 1),
  unit = NULL,
  breaks,
  logscale = TRUE
) {
  if (logscale == TRUE) {
    values <- scales::rescale(scales::pseudo_log_trans()$transform(breaks))
  } else {
    values <- scales::rescale(breaks)
  }

  ggplot() +
    geom_sf(
      data = shift_geometry(dat),
      aes(fill = get(var)),
      color = NA
    ) +
    geom_sf(data = shift_geometry(states_pop), fill = NA, color = 'grey25') +
    scale_fill_stepsn(
      colors = khroma::color('incandescent')(5),
      name = str_wrap(
        if_else(is.null(unit), var, paste0(var, '(', unit, ')')),
        width = 12
      ),
      labels = flab,
      n.breaks = 5,
      breaks = breaks,
      values = values,
      transform = if_else(
        logscale == TRUE,
        'pseudo_log',
        'identity'
      )
    ) +
    coord_sf(expand = FALSE) +
    theme(
      axis.ticks.x = element_blank(),
      axis.ticks.y = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_blank(),
      axis.line = element_blank(),
      plot.margin = margin(0, 0, 0, 0),
      legend.margin = margin(0, 0, 0, l = 75),
      #legend.position = 'bottom',
      legend.position = "left",
      legend.justification = "top",
      legend.text = element_text(hjust = 0, vjust = -.5),
      panel.spacing = unit(0, "lines")
    )
}
```

### Number of Fire Disasters per Region over Time

```{r events_over_time, fig.height = 12, fig.width = 8}
plotme(wfbz_crit, 'Events', rel = FALSE) +
  plot_spacer() +
  plotme(wfbz_crit, 'Events', rel = TRUE) +
  plot_spacer() +
  free(
    mapme(
      wfbz_hex,
      'Events',
      breaks = exp(pretty(log(range(wfbz_hex$Events, na.rm = TRUE)), n = 5))
    ),
    type = 'panel'
  ) +
  plot_layout(ncol = 1, heights = c(5, 1, 5, .5, 6), widths = c(3, 3, 3, 3, 4))
ggsave(
  here('figures/events_over_time.pdf'),
  last_plot(),
  width = 8,
  height = 12
)
```


### Number of Fatalities per Region over Time

The five deadliest fires in our data set are labeled.


```{r fatalities_over_time, fig.height = 12, fig.width = 8}
plotme(wfbz_crit, 'Fatalities', rel = FALSE) +
  annotate(
    "text",
    x = 2023,
    y = 50,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Lahaina Fire (HI) ———"
  ) +
  annotate(
    "text",
    x = 2018,
    y = 80,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Camp Fire (CA) ———"
  ) +
  annotate(
    "text",
    x = 2017,
    y = 50,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Tubbs Fire (CA) ———"
  ) +
  annotate(
    "text",
    x = 2013,
    y = 19,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Yarnell Hill Fire (AZ) ———"
  ) +
  annotate(
    "text",
    x = 2020,
    y = 20,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "North Complex Fire (CA) ———"
  ) +
  plot_spacer() +
  plotme(wfbz_crit, 'Fatalities', rel = TRUE) +
  plot_spacer() +
  free(
    mapme(
      wfbz_hex,
      'Fatalities',
      flab = ~ scales::comma(.x, accuracy = .001),
      breaks = c(
        0,
        scales::breaks_log(n = 5, base = 10)(c(
          0.001,
          max(wfbz_hex$`Fatalities per 100K`, na.rm = TRUE)
        ))
      ),
    ),

    type = 'panel'
  ) +
  plot_layout(ncol = 1, heights = c(5, 1, 5, .5, 6), widths = c(3, 3, 3, 3, 4))
ggsave(
  here('figures/fatalities_over_time.pdf'),
  last_plot(),
  width = 8,
  height = 12
)
```


### Number of Structures Destroyed per Region over Time

The five fires in our data set with the most destroyed structures are labeled.

```{r structures_over_time, fig.height = 12, fig.width = 8}
plotme(wfbz_crit, 'Structures Destroyed', rel = FALSE) +
  annotate(
    "text",
    x = 2018,
    y = 154 * 100,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Camp Fire (CA) ———"
  ) +
  annotate(
    "text",
    x = 2017,
    y = 85 * 100,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Tubbs Fire (CA) ———"
  ) +
  annotate(
    "text",
    x = 2023,
    y = 22 * 100,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Lahaina Fire (HI) ———"
  ) +
  annotate(
    "text",
    x = 2015,
    y = 25 * 100,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Valley Fire (CA) ———"
  ) +
  annotate(
    "text",
    x = 2018,
    y = 140 * 100,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Woolsey (CA) ———"
  ) +
  plot_spacer() +
  plotme(wfbz_crit, 'Structures Destroyed', rel = TRUE) +
  plot_spacer() +
  free(
    mapme(
      wfbz_hex,
      'Structures Destroyed',
      breaks = c(
        0,
        scales::breaks_log(n = 6, base = 10)(c(
          1,
          max(wfbz_hex$`Structures Destroyed`, na.rm = TRUE)
        ))
      )
    ),
    type = 'panel'
  ) +
  plot_layout(ncol = 1, heights = c(5, 1, 5, .5, 6), widths = c(3, 3, 3, 3, 4))
ggsave(
  here('figures/structures_over_time.pdf'),
  last_plot(),
  width = 8,
  height = 12
)
```


### Number of FEMA Declarations per Region over Time

```{r fema_over_time, fig.height = 12, fig.width = 8}
plotme(wfbz_crit, 'FEMA Declarations (n)', rel = FALSE) +
  plot_spacer() +
  plotme(wfbz_crit, 'FEMA Declarations (n)', rel = TRUE) +
  plot_spacer() +
  free(
    mapme(
      wfbz_hex,
      'FEMA Declarations (n)',
      breaks = c(
        0,
        scales::breaks_log(n = 6, base = 10)(c(
          1,
          max(wfbz_hex$`FEMA Declarations (n)`, na.rm = TRUE)
        ))
      ),
    ),
    type = 'panel'
  ) +
  plot_layout(ncol = 1, heights = c(5, 1, 5, .5, 6), widths = c(3, 3, 3, 3, 4))
ggsave(
  here('figures/fema_over_time.pdf'),
  last_plot(),
  width = 8,
  height = 12
)
```

### Burned Area over Time

_i.e. the total area of all wildfire footprints per region/year (even if a fire affected the same spot twice in the year)_

```{r burned_area_over_time, fig.height = 12, fig.width = 8}
plotme(wfbz_crit, 'Burned Area', rel = FALSE, unit = 'km²') +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0)),
    labels = scales::comma
  ) +
  plot_spacer() +
  plotme(wfbz_crit, 'Burned Area', rel = TRUE) +
  plot_spacer() +
  free(
    mapme(
      wfbz_hex,
      'Burned Area',
      unit = 'km²',
      breaks = c(
        0,
        scales::breaks_log(n = 6, base = 10)(c(
          1,
          max(wfbz_hex$`Burned Area`, na.rm = TRUE)
        ))
      )
    ),
    type = 'panel'
  ) +
  plot_layout(ncol = 1, heights = c(5, 1, 5, .5, 6), widths = c(3, 3, 3, 3, 4))
ggsave(
  here('figures/burned_area_over_time.pdf'),
  last_plot(),
  width = 8,
  height = 12
)
```

### Average Fire Size over Time

_i.e. the mean area of the wildfire footprints in a region/year_


```{r fire_size_over_time, fig.height = 8, fig.width = 8}
plotme(wfbz_crit, 'Mean Fire Size', rel = FALSE, unit = 'km²') +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0)),
    labels = scales::comma
  ) +
  plot_spacer() +
  free(
    mapme(
      wfbz_hex,
      'Mean Fire Size',
      unit = 'km²',
      breaks = c(
        0,
        scales::breaks_log(n = 6, base = 10)(c(
          1,
          max(wfbz_hex$`Mean Fire Size`, na.rm = TRUE)
        ))
      )
    ),
    type = 'panel'
  ) +
  plot_layout(ncol = 1, heights = c(5, .5, 6), widths = c(3, 3, 4))
ggsave(
  here('figures/fire_size_over_time.pdf'),
  last_plot(),
  width = 8,
  height = 8
)
```

## Share of WUI Fires over Time

```{r wui_over_time, fig.height = 10, fig.width = 14}
plotme(wfbz_crit, 'WUI', stack = FALSE) +
  xlab('') +
  theme(
    margins = margin(10, 10, 10, 10, unit = 'pt'),
    panel.spacing = unit(10, "pt")
  )
ggsave(
  here('figures/wui_over_time.pdf'),
  last_plot(),
  width = 10,
  height = 16
)
```

```{r wui_over_time_map, fig.height = 4, fig.width = 8}
mapme(
  wfbz_hex,
  'WUI Percent',
  breaks = seq(0, 100, by = 20),
  flab = ~ paste0(round(.x), '%'),
  logscale = FALSE
)
ggsave(
  here('figures/wui_over_time_map.pdf'),
  last_plot(),
  width = 14,
  height = 6
)
```
