---
title: "Variables over Time"
toc: true
execute:
  echo: false
  warning: false
  message: false
---

{{< include _setup.qmd >}}

```{r map_setup}
hexgrid <- st_make_grid(states_pop, square = FALSE, n = c(500, 500)) %>%
  st_as_sf() %>%
  mutate(hexid = row_number()) %>%
  st_filter(states_pop)


wfbz_hex <- hexgrid %>%
  st_join(
    wfbz %>%
      mutate(
        usfs_region = wfbz_region,
        usfs_region_area = wfbz_region_area,
        usfs_region_pop = wfbz_region_pop
      ),
    left = FALSE
  ) %>%
  filter(wildfire_community_intersect) %>%
  group_by(hexid, geometry = x) %>%
  summarize(
    `Events` = n(),
    `Fatalities` = sum(wildfire_max_civil_fatalities, na.rm = TRUE),
    `Fatalities per 100K` = 100000 *
      sum(wildfire_max_civil_fatalities / usfs_region_pop, na.rm = TRUE),
    `Structures Destroyed` = sum(wildfire_struct_destroyed, na.rm = TRUE),
    `FEMA Declarations (n)` = sum(wildfire_fema_dec, na.rm = TRUE),
    `Burned Area` = sum(wildfire_area, na.rm = TRUE),
    `Burned Area Percent` = 100 *
      sum(wildfire_area / usfs_region_area, na.rm = TRUE),
    `WUI` = sum(str_detect(wildfire_wui, '(intermix|interface)'), na.rm = TRUE),
    `WUI Percent` = 100 *
      sum(str_detect(wildfire_wui, '(interface|intermix)'), na.rm = TRUE) /
      n(),
    `Mean Fire Size` = mean(wildfire_area, na.rm = TRUE),
    .groups = 'drop'
  )

mapme <- function(dat, var, flab = scales::comma, unit = NULL) {
  ggplot() +
    geom_sf(
      data = shift_geometry(dat),
      aes(fill = get(var)),
      color = NA
    ) +
    geom_sf(data = shift_geometry(states_pop), fill = NA, color = 'grey25') +
    khroma::scale_fill_iridescent(
      name = if_else(is.null(unit), var, paste0(var, '(', unit, ')')),
      labels = flab
    ) +
    coord_sf(expand = FALSE) +
    theme(
      axis.ticks.x = element_blank(),
      axis.ticks.y = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_blank(),
      axis.line = element_blank(),
      plot.margin = margin(0, 0, 0, 0)
    )
}
```

### Number of Fire Disasters per Region over Time

```{r events_over_time, fig.height = 12, fig.width = 8}
plotme(wfbz_crit, 'Events', rel = FALSE) +
  plot_spacer() +
  plotme(wfbz_crit, 'Events', rel = TRUE) +
  plot_spacer() +
  mapme(wfbz_hex, 'Events') +
  plot_layout(ncol = 1, heights = c(5, 1, 5, 1, 5))
ggsave(
  here('figures/events_over_time.pdf'),
  last_plot(),
  width = 8,
  height = 12
)
```


### Number of Fatalities per Region over Time

The five deadliest fires in our data set are labeled.


```{r fatalities_over_time, fig.height = 12, fig.width = 8}
plotme(wfbz_crit, 'Fatalities', rel = FALSE) +
  annotate(
    "text",
    x = 2023,
    y = 50,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Lahaina Fire (HI) ———"
  ) +
  annotate(
    "text",
    x = 2018,
    y = 80,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Camp Fire (CA) ———"
  ) +
  annotate(
    "text",
    x = 2017,
    y = 50,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Tubbs Fire (CA) ———"
  ) +
  annotate(
    "text",
    x = 2013,
    y = 19,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Yarnell Hill Fire (AZ) ———"
  ) +
  annotate(
    "text",
    x = 2020,
    y = 20,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "North Complex Fire (CA) ———"
  ) +
  plot_spacer() +
  plotme(wfbz_crit, 'Fatalities', rel = TRUE) +
  plot_spacer() +
  mapme(wfbz_hex, 'Fatalities') +
  plot_layout(ncol = 1, heights = c(5, 1, 5, 1, 5))
ggsave(
  here('figures/fatalities_over_time.pdf'),
  last_plot(),
  width = 8,
  height = 12
)
```


### Number of Structures Destroyed per Region over Time

The five fires in our data set with the most destroyed structures are labeled.

```{r structures_over_time, fig.height = 12, fig.width = 8}
plotme(wfbz_crit, 'Structures Destroyed', rel = FALSE) +
  annotate(
    "text",
    x = 2018,
    y = 154 * 100,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Camp Fire (CA) ———"
  ) +
  annotate(
    "text",
    x = 2017,
    y = 85 * 100,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Tubbs Fire (CA) ———"
  ) +
  annotate(
    "text",
    x = 2023,
    y = 22 * 100,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Lahaina Fire (HI) ———"
  ) +
  annotate(
    "text",
    x = 2015,
    y = 25 * 100,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Valley Fire (CA) ———"
  ) +
  annotate(
    "text",
    x = 2018,
    y = 140 * 100,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Woolsey (CA) ———"
  ) +
  plot_spacer() +
  plotme(wfbz_crit, 'Structures Destroyed', rel = TRUE) +
  plot_spacer() +
  mapme(wfbz_hex, 'Structures Destroyed') +
  plot_layout(ncol = 1, heights = c(5, 1, 5, 1, 5))
ggsave(
  here('figures/structures_over_time.pdf'),
  last_plot(),
  width = 8,
  height = 12
)
```


### Number of FEMA Declarations per Region over Time

```{r fema_over_time, fig.height = 12, fig.width = 8}
plotme(wfbz_crit, 'FEMA Declarations (n)', rel = FALSE) +
  plot_spacer() +
  plotme(wfbz_crit, 'FEMA Declarations (n)', rel = TRUE) +
  plot_spacer() +
  mapme(wfbz_hex, 'FEMA Declarations (n)') +
  plot_layout(ncol = 1, heights = c(5, 1, 5, 1, 5))
ggsave(
  here('figures/fema_over_time.pdf'),
  last_plot(),
  width = 8,
  height = 12
)
```

### Burned Area over Time

_i.e. the total area of all wildfire footprints per region/year (even if a fire affected the same spot twice in the year)_

```{r burned_area_over_time, fig.height = 12, fig.width = 8}
plotme(wfbz_crit, 'Burned Area', rel = FALSE, unit = 'km²') +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0)),
    labels = scales::comma
  ) +
  plot_spacer() +
  plotme(wfbz_crit, 'Burned Area', rel = TRUE) +
  plot_spacer() +
  mapme(wfbz_hex, 'Burned Area', unit = 'km²') +
  plot_layout(ncol = 1, heights = c(5, 1, 5, 1, 5))
ggsave(
  here('figures/burned_area_over_time.pdf'),
  last_plot(),
  width = 8,
  height = 12
)
```

### Average Fire Size over Time

_i.e. the mean area of the wildfire footprints in a region/year_


```{r fire_size_over_time, fig.height = 8, fig.width = 8}
plotme(wfbz_crit, 'Mean Fire Size', rel = FALSE, unit = 'km²') +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0)),
    labels = scales::comma
  ) +
  plot_spacer() +
  mapme(wfbz_hex, 'Mean Fire Size', unit = 'km²') +
  plot_layout(ncol = 1, heights = c(5, 1, 5))
ggsave(
  here('figures/fire_size_over_time.pdf'),
  last_plot(),
  width = 8,
  height = 8
)
```

## Share of WUI Fires over Time

```{r wui_over_time, fig.height = 10, fig.width = 14}
plotme(wfbz_crit, 'WUI', stack = FALSE) +
  xlab('') +
  theme(
    margins = margin(10, 10, 10, 10, unit = 'pt'),
    panel.spacing = unit(10, "pt")
  )
ggsave(
  here('figures/wui_over_time.pdf'),
  last_plot(),
  width = 10,
  height = 16
)
```

```{r wui_over_time_map, fig.height = 4, fig.width = 8}
mapme(wfbz_hex, 'WUI Percent', flab = \(x) scales::percent(x / 100))
ggsave(
  here('figures/wui_over_time_map.pdf'),
  last_plot(),
  width = 14,
  height = 6
)
```
