---
title: "Variables over Time"
toc: true
execute:
  echo: false
  warning: false
  message: false
---

{{< include _setup.qmd >}}

### Number of Fire Disasters per Region over Time

```{r events_over_time, fig.height = 8, fig.width = 16}
plotme(wfbz_crit, 'Events', rel = FALSE) +
  ggtitle('Community-Impacting Disasters') +
  plotme(wfbz_crit_all, 'Events', rel = FALSE) +
  ggtitle('All Disasters') +
  plot_spacer() +
  plot_spacer() +
  plotme(wfbz_crit, 'Events', rel = TRUE) +
  plotme(wfbz_crit_all, 'Events', rel = TRUE) +
  plot_layout(ncol = 2, heights = c(5, 1, 5))
ggsave(here('figures/events_over_time.png'), last_plot(), width = 16, height = 8)
```


### Number of Fatalities per Region over Time

The five deadliest fires in our data set are labeled.


```{r fatalities_over_time, fig.height = 8, fig.width = 16}
(plotme(wfbz_crit, 'Fatalities', rel = FALSE) +
  annotate(
    "text",
    x = 2023,
    y = 50,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Lahaina Fire (HI) ———"
  ) +
  annotate(
    "text",
    x = 2018,
    y = 80,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Camp Fire (CA) ———"
  ) +
  annotate(
    "text",
    x = 2017,
    y = 50,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Tubbs Fire (CA) ———"
  ) +
  annotate(
    "text",
    x = 2013,
    y = 19,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Yarnell Hill Fire (AZ) ———"
  ) +
  annotate(
    "text",
    x = 2020,
    y = 20,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "North Complex Fire (CA) ———"
  ) +
  ggtitle('Community-Impacting Disasters')) +
  (plotme(wfbz_crit_all, 'Fatalities', rel = FALSE) +
    ggtitle('All Disasters') +
    annotate(
      "text",
      x = 2023,
      y = 50,
      size = 2,
      angle = -45,
      hjust = 'right',
      label = "Lahaina Fire (HI) ———"
    ) +
    annotate(
      "text",
      x = 2018,
      y = 80,
      size = 2,
      angle = -45,
      hjust = 'right',
      label = "Camp Fire (CA) ———"
    ) +
    annotate(
      "text",
      x = 2017,
      y = 50,
      size = 2,
      angle = -45,
      hjust = 'right',
      label = "Tubbs Fire (CA) ———"
    ) +
    annotate(
      "text",
      x = 2013,
      y = 19,
      size = 2,
      angle = -45,
      hjust = 'right',
      label = "Yarnell Hill Fire (AZ) ———"
    ) +
    annotate(
      "text",
      x = 2020,
      y = 20,
      size = 2,
      angle = -45,
      hjust = 'right',
      label = "North Complex Fire (CA) ———"
    )) +
  plot_spacer() +
  plot_spacer() +
  plotme(wfbz_crit, 'Fatalities', rel = TRUE) +
  plotme(wfbz_crit_all, 'Fatalities', rel = TRUE) +
  plot_layout(ncol = 2, heights = c(5, 1, 5))
ggsave(here('figures/fatalities_over_time.png'), last_plot(), width = 16, height = 8)
```


### Number of Structures Destroyed per Region over Time

The five fires in our data set with the most destroyed structures are labeled.

```{r structures_over_time, fig.height = 8, fig.width = 16}
(plotme(wfbz_crit, 'Structures Destroyed', rel = FALSE) +
  annotate(
    "text",
    x = 2018,
    y = 154 * 100,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Camp Fire (CA) ———"
  ) +
  annotate(
    "text",
    x = 2017,
    y = 85 * 100,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Tubbs Fire (CA) ———"
  ) +
  annotate(
    "text",
    x = 2023,
    y = 22 * 100,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Lahaina Fire (HI) ———"
  ) +
  annotate(
    "text",
    x = 2015,
    y = 25 * 100,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Valley Fire (CA) ———"
  ) +
  annotate(
    "text",
    x = 2018,
    y = 140 * 100,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Woolsey (CA) ———"
  ) +
  ggtitle('Community-Impacting Disasters')) +
  (plotme(wfbz_crit_all, 'Structures Destroyed', rel = FALSE) +
    ggtitle('All Disasters') +
    annotate(
    "text",
    x = 2018,
    y = 154 * 100,
    size = 2,
    angle = -45,
    hjust = 'right',
    label = "Camp Fire (CA) ———"
    ) +
    annotate(
      "text",
      x = 2017,
      y = 85 * 100,
      size = 2,
      angle = -45,
      hjust = 'right',
      label = "Tubbs Fire (CA) ———"
    ) +
    annotate(
      "text",
      x = 2023,
      y = 22 * 100,
      size = 2,
      angle = -45,
      hjust = 'right',
      label = "Lahaina Fire (HI) ———"
    ) +
    annotate(
      "text",
      x = 2015,
      y = 25 * 100,
      size = 2,
      angle = -45,
      hjust = 'right',
      label = "Valley Fire (CA) ———"
    ) +
    annotate(
      "text",
      x = 2018,
      y = 140 * 100,
      size = 2,
      angle = -45,
      hjust = 'right',
      label = "Woolsey (CA) ———"
    )) +
  plot_spacer() +
  plot_spacer() +
  plotme(wfbz_crit, 'Structures Destroyed', rel = TRUE) +
  plotme(wfbz_crit_all, 'Structures Destroyed', rel = TRUE) +
  plot_layout(ncol = 2, heights = c(5, 1, 5))
ggsave(here('figures/structures_over_time.png'), last_plot(), width = 16, height = 8)
```


### Number of FEMA Declarations per Region over Time

```{r fema_over_time, fig.height = 8, fig.width = 16}
plotme(wfbz_crit, 'FEMA Declarations (n)', rel = FALSE) +
  ggtitle('Community-Impacting Disasters') +
  plotme(wfbz_crit_all, 'FEMA Declarations (n)', rel = FALSE) +
  ggtitle('All Disasters') +
  plot_spacer() +
  plot_spacer() +
  plotme(wfbz_crit, 'FEMA Declarations (n)', rel = TRUE) +
  plotme(wfbz_crit_all, 'FEMA Declarations (n)', rel = TRUE) +
  plot_layout(ncol = 2, heights = c(5, 1, 5))
ggsave(here('figures/fema_over_time.png'), last_plot(), width = 16, height = 8)
```

### Burned Area over Time

```{r burned_area_over_time, fig.height = 8, fig.width = 16}
(plotme(wfbz_crit, 'Burned Area', rel = FALSE, unit = 'km²') +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0)), labels = scales::comma
  ) +
  ggtitle('Community-Impacting Disasters')) +
  (plotme(wfbz_crit_all, 'Burned Area', rel = FALSE, unit = 'km²') +
    #scale_y_continuous(expand = expansion(mult = c(0, 0)), labels = scales::scientific) +
    ggtitle('All Disasters')) +
  plot_spacer() +
  plot_spacer() +
  plotme(wfbz_crit, 'Burned Area', rel = TRUE) +
  plotme(wfbz_crit_all, 'Burned Area', rel = TRUE) +
  plot_layout(ncol = 2, heights = c(5, 1, 5))
ggsave(here('figures/burned_area_over_time.png'), last_plot(), width = 16, height = 8)
```

### Average Fire Size over Time

```{r fire_size_over_time, fig.height = 4, fig.width = 16}
(plotme(wfbz_crit, 'Mean Fire Size', rel = FALSE, unit = 'km²') +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0)),
    labels = scales::comma
  ) +
  ggtitle('Community-Impacting Disasters')) +
  (plotme(wfbz_crit_all, 'Mean Fire Size', rel = FALSE, unit = 'km²') +
    #scale_y_continuous(expand = expansion(mult = c(0, 0)), labels = scales::scientific) +
    ggtitle('All Disasters')) +
  plot_layout(ncol = 2)
ggsave(here('figures/fire_size_over_time.png'), last_plot(), width = 16, height = 4)
```


Alaska has quite an outlier for community-intersecting fires in 2004. This was the year of the Boundary Fire NE of Fairbanks, which reached 912 km<sup>2</sup>. While that's not the largest fire in our data set, it was the _only_ community-intersecting fire in our data set that year in Alaska.

## Share of WUI Fires over Time

### Share of Fires in Wildland-Urban Interface over Time

```{r wui_interface_over_time, fig.height = 12, fig.width = 14}
plotme(wfbz_crit, 'WUI-Interface', stack = FALSE) +
  ggtitle('Community-Impacting Disasters') +
  xlab('')
ggsave(here('figures/wui_interface_over_time.png'), last_plot(), width = 14, height = 12)
```

### Share of Fires in Wildland-Urban Intermix over Time

```{r wui_intermix_over_time, fig.height = 12, fig.width = 14}
plotme(wfbz_crit, 'WUI-Intermix', stack = FALSE) +
  ggtitle('Community-Impacting Disasters') +
  xlab('')
ggsave(here('figures/wui_intermix_over_time.png'), last_plot(), width = 14, height = 12)
```
