---
title: "Demographic Time Series"
toc: true
execute:
  echo: false
  warning: false
  message: false
---

{{< include _pop_setup.qmd >}}

In each plot, the colored, solid line represents the median value for the metric across wildfire-affected census tracts.
The colored ribbon ranges from the 25th to the 75th percentile for that metric.

The same is show for "at risk" census tracts (the 90th percentile of tracts nationally by WHP) in grey, with the dashed
line showing the median.

```{r load_data}
# Fire-tract census joins (built by staging/prep_spatial_join.R)
by_wf_dir <- here('data/processed/nhgis')
race_by_wf <- read_parquet(file.path(by_wf_dir, 'race_by_wf.parquet'))
poverty_by_wf <- read_parquet(file.path(by_wf_dir, 'poverty_by_wf.parquet'))
vehicle_avail_by_wf <- read_parquet(file.path(
  by_wf_dir,
  'vehicle_avail_by_wf.parquet'
))
housing_cost_by_wf <- read_parquet(file.path(
  by_wf_dir,
  'housing_cost_by_wf.parquet'
))

# Regional population reference fracs
region_pop_vehicle <- read_parquet(here(
  'data/processed/nhgis/region_pop_vehicle.parquet'
))
region_pop_race <- read_parquet(here(
  'data/processed/nhgis/region_pop_race.parquet'
))
region_pop_poverty <- read_parquet(here(
  'data/processed/nhgis/region_pop_poverty.parquet'
))
region_pop_housing <- read_parquet(here(
  'data/processed/nhgis/region_pop_housing.parquet'
))

# Region assignment for each fire
wfbz <- read_sf(here('data/raw/wfbz_disasters_2000-2025.geojson')) %>%
  filter(wildfire_year < 2025) %>%
  filter(wildfire_community_intersect)

regions <- read_sf(here('data/processed/regions.geojson'))
wfbz_region <- wfbz %>%
  filter(!st_is_empty(geometry)) %>%
  select(wildfire_id) %>%
  mutate(
    usfs_region = regions$usfs_region[st_nearest_feature(geometry, regions)]
  ) %>%
  st_drop_geometry()
```

```{r plot_fn}
plot_pop_var <- function(df, label) {
  df %>%
    ggplot() +
    # Ref
    geom_ribbon(
      aes(
        x = wildfire_year,
        ymin = frac_pop_p25,
        ymax = frac_pop_p75
      ),
      fill = 'grey50',
      color = 'grey25',
      linewidth = 0.25,
      alpha = .05
    ) +
    geom_line(
      aes(x = wildfire_year, y = frac_pop_median, color = 'grey25'),
      linetype = 'dashed',
      linewidth = 0.4
    ) +
    geom_point(
      aes(x = wildfire_year, y = frac_pop_median, color = 'grey25'),
      size = 1,
      shape = 1
    ) +
    # Affected
    geom_line(aes(x = wildfire_year, y = frac_median, color = usfs_region)) +
    geom_point(
      aes(x = wildfire_year, y = frac_median, color = usfs_region),
      size = 1
    ) +
    geom_ribbon(
      aes(
        x = wildfire_year,
        ymin = frac_p25,
        ymax = frac_p75,
        fill = usfs_region,
        color = usfs_region
      ),
      alpha = .5
    ) +
    scale_color_manual(name = "Region", values = region_colors) +
    scale_fill_manual(name = "Region", values = region_colors) +
    scale_x_continuous(
      breaks = seq(2000, 2024, by = 2),
      expand = expansion(add = c(.25, .25))
    ) +
    coord_cartesian(xlim = c(2000, 2024)) +
    scale_y_continuous(labels = scales::percent) +
    facet_wrap(~usfs_region, ncol = 1, scales = 'free_y') +
    xlab(NULL) +
    ylab(label) +
    pop_theme
}
```

## Vehicle Availability

```{r vehicle_avail_fig, fig.height = 16}
#| fig-cap: "Share of fire-affected households with no vehicle access, by USFS region and year"
vehicle_avail_by_region <- vehicle_avail_by_wf %>%
  inner_join(wfbz_region, by = 'wildfire_id') %>%
  group_by(usfs_region, wildfire_year, wildfire_id) %>%
  summarize(
    pct_no_vehicle = sum(count[no_vehicle == TRUE], na.rm = TRUE) /
      sum(count, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  group_by(wildfire_year, usfs_region) %>%
  summarize(
    frac_p25 = quantile(pct_no_vehicle, .25, na.rm = TRUE),
    frac_median = median(pct_no_vehicle, na.rm = TRUE),
    frac_p75 = quantile(pct_no_vehicle, .75, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  left_join(region_pop_vehicle, by = c('wildfire_year', 'usfs_region')) %>%
  inner_join(REGION_ORDER, by = 'usfs_region') %>%
  mutate(usfs_region = fct_reorder(usfs_region, plot_order))

plot_pop_var(vehicle_avail_by_region, "Share with no vehicle")
ggsave(
  here('figures/vehicle_avail_fig.pdf'),
  last_plot(),
  width = 7,
  height = 16
)
```

## Race

```{r race_fig, fig.height = 16}
#| fig-cap: "Share of fire-affected population that is Non-Hispanic White, by USFS region and year"
race_by_region <- race_by_wf %>%
  inner_join(wfbz_region, by = 'wildfire_id') %>%
  group_by(usfs_region, wildfire_year, wildfire_id) %>%
  summarize(
    pct_nhw = sum(
      count[
        ethnicity == "Not Hispanic or Latino" & race == "White (single race)"
      ],
      na.rm = TRUE
    ) /
      sum(count, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  group_by(wildfire_year, usfs_region) %>%
  summarize(
    frac_p25 = quantile(pct_nhw, .25, na.rm = TRUE),
    frac_median = median(pct_nhw, na.rm = TRUE),
    frac_p75 = quantile(pct_nhw, .75, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  left_join(region_pop_race, by = c('wildfire_year', 'usfs_region')) %>%
  inner_join(REGION_ORDER, by = 'usfs_region') %>%
  mutate(usfs_region = fct_reorder(usfs_region, plot_order))

plot_pop_var(race_by_region, "Share Non-Hispanic White")
ggsave(here('figures/race_fig.pdf'), last_plot(), width = 7, height = 16)
```

## Poverty

```{r poverty_fig, fig.height = 16}
#| fig-cap: "Population below 100% of poverty line near wildfire burn zones, by USFS region and year"
poverty_by_region <- poverty_by_wf %>%
  inner_join(wfbz_region, by = 'wildfire_id') %>%
  group_by(usfs_region, wildfire_year, wildfire_id) %>%
  summarize(
    pct_poverty = sum(count[poverty_lt_100 == TRUE], na.rm = TRUE) /
      sum(count, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  group_by(wildfire_year, usfs_region) %>%
  summarize(
    frac_p25 = quantile(pct_poverty, .25, na.rm = TRUE),
    frac_median = median(pct_poverty, na.rm = TRUE),
    frac_p75 = quantile(pct_poverty, .75, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  left_join(region_pop_poverty, by = c('wildfire_year', 'usfs_region')) %>%
  inner_join(REGION_ORDER, by = 'usfs_region') %>%
  mutate(usfs_region = fct_reorder(usfs_region, plot_order))

plot_pop_var(poverty_by_region, "Share below 100% poverty line")
ggsave(here('figures/poverty_fig.pdf'), last_plot(), width = 7, height = 16)
```

## Housing Cost Burden

```{r housing_cost_fig, fig.height = 16}
#| fig-cap: "Cost-burdened households (>=50% income on housing) near wildfire burn zones, by USFS region and year"
housing_cost_by_region <- housing_cost_by_wf %>%
  inner_join(wfbz_region, by = 'wildfire_id') %>%
  group_by(usfs_region, wildfire_year, wildfire_id) %>%
  summarize(
    pct_cost_burdened = sum(count[cost_burdened == TRUE], na.rm = TRUE) /
      sum(count, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  group_by(wildfire_year, usfs_region) %>%
  summarize(
    frac_p25 = quantile(pct_cost_burdened, .25, na.rm = TRUE),
    frac_median = median(pct_cost_burdened, na.rm = TRUE),
    frac_p75 = quantile(pct_cost_burdened, .75, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  left_join(region_pop_housing, by = c('wildfire_year', 'usfs_region')) %>%
  inner_join(REGION_ORDER, by = 'usfs_region') %>%
  mutate(usfs_region = fct_reorder(usfs_region, plot_order))

plot_pop_var(housing_cost_by_region, "Share housing cost burdened (>=50%)")
ggsave(
  here('figures/housing_cost_fig.pdf'),
  last_plot(),
  width = 7,
  height = 16
)
```
