---
title: "WHP Diagnostics"
toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r load_data}
library(sf)
library(tidyverse)
library(here)
library(tigris)
library(arrow)
tract_sf <- bind_rows(
  read_sf(here('data/processed/tract_WHP_AK.geojson')) %>%
    filter(substr(GEOID, 1, 2) == '02'),
  read_sf(here('data/processed/tract_WHP_HI.geojson')) %>%
    filter(substr(GEOID, 1, 2) == '15'),
  read_sf(here('data/processed/tract_WHP_CONUS.geojson')) %>%
    filter(!(substr(GEOID, 1, 2) %in% c('02', '15')))
) %>%
  filter(census_year == 2020) %>%
  mutate(
    high_risk = whp_mean >= quantile(whp_mean, 0.9, na.rm = TRUE),
    high_risk_75 = whp_mean >= 75 #quantile(whp_mean, 0.67, na.rm = TRUE)
  ) %>%
  left_join(
    read_parquet(here('data/processed/nhgis/population.parquet')) %>%
      filter(period == '2019-2023')
  )
```

```{r}
county_sf <- tract_sf %>%
  group_by(county = substr(GEOID, 1, 5)) %>%
  summarize(
    whp_mean = mean(whp_mean, na.rm = TRUE),
    whp_max = max(whp_mean, na.rm = TRUE),
    pct_high_risk = mean(high_risk),
    pct_high_risk_75 = mean(high_risk_75),
    geometry = st_union(geometry)
  )
```

```{r}
wfbz <- read_sf(here('data/raw/wfbz_disasters_2000-2025.geojson'))
```

### WHP Descriptives

(by tract)

```{r}
summary(tract_sf$whp_mean)
```

```{r}
ggplot(tract_sf) + geom_density(aes(whp_mean)) + scale_x_log10()
```

### WHP Risk cutoff

```{r}
quantile(tract_sf$whp_mean, 0.90, na.rm = TRUE)
```

```{r}
#quantile(tract_sf$whp_mean, 0.67, na.rm = TRUE)
75
```


### pop included

top decile

```{r}
sum(tract_sf$high_risk * tract_sf$pop, na.rm = T) /
  sum(tract_sf$pop, na.rm = T)
```

mean value at thrshold

```{r}
sum(tract_sf$high_risk_75 * tract_sf$pop, na.rm = T) /
  sum(tract_sf$pop, na.rm = T)
```


## WHP Distribution

### County

#### Mean tract-WHP

natural scale

```{r}
ggplot(shift_geometry(county_sf)) +
  geom_sf(aes(fill = whp_mean), color = NA, linewidth = 0) +
  scale_fill_fermenter(
    palette = 'Spectral',
    direction = -1,
    breaks = ~ quantile(.x, seq(.1, .9, by = .1))
  ) +
  theme_void()
```


#### Max tract-WHP

(log transformed)

```{r}
ggplot(shift_geometry(county_sf)) +
  geom_sf(aes(fill = whp_max), color = NA, linewidth = 0) +
  scale_fill_fermenter(
    palette = 'Spectral',
    direction = -1,
    transform = 'log',
    labels = scales::comma,
    n.breaks = 8
  ) +
  theme_void()
```

### High risk

top decile

```{r}
ggplot(shift_geometry(county_sf)) +
  geom_sf(aes(fill = pct_high_risk), color = NA, linewidth = 0) +
  scale_fill_fermenter(
    palette = 'Spectral',
    direction = -1,
    labels = scales::percent,
    breaks = ~ quantile(.x, seq(.1, .9, by = .1))
  ) +
  theme_void()
```

mean value at thrshold

```{r}
ggplot(shift_geometry(county_sf)) +
  geom_sf(aes(fill = pct_high_risk_75), color = NA, linewidth = 0) +
  scale_fill_fermenter(
    palette = 'Spectral',
    direction = -1,
    labels = scales::percent,
    breaks = ~ quantile(.x, seq(.1, .9, by = .1))
  ) +
  theme_void()
```

## WFBZs vs Risk

```{r}
wfbz_whp <- wfbz %>%
  st_join(st_transform(tract_sf, st_crs(wfbz))) %>%
  st_drop_geometry() %>%
  group_by(wildfire_id) %>%
  summarize(
    whp_min = min(whp_mean, na.rm = TRUE),
    whp_mean = mean(whp_mean, na.rm = TRUE),
    whp_max = max(whp_mean, na.rm = TRUE),
    any_high_risk = any(high_risk, na.rm = TRUE),
    pct_high_risk = mean(high_risk, na.rm = TRUE),
    any_high_risk_75 = any(high_risk_75, na.rm = TRUE),
    pct_high_risk_75 = mean(high_risk_75, na.rm = TRUE)
  )
```

### Fires where _any_ tract that burned was high risk:

Top decile

```{r}
table(wfbz_whp$any_high_risk)
```

mean value at thrshold
```{r}
table(wfbz_whp$any_high_risk_75)
```

### Percent of burned tracts that were high risk (by fire)

Top decile

```{r}
summary(wfbz_whp$pct_high_risk)
```

pct at 100%

```{r}
mean(wfbz_whp$pct_high_risk == 1, na.rm = T)
```

mean value at thrshold

```{r}
summary(wfbz_whp$pct_high_risk_75)
```

pct at 100%
```{r}
mean(wfbz_whp$pct_high_risk_75 == 1, na.rm = T)
```
