---
title: "WUI Tables"
toc: true
execute:
  echo: false
  warning: false
  message: false
---

{{< include _setup.qmd >}}

```{r}
pop_affected_annual <- bind_rows(
  'WUI' = read_csv(here('data/processed/pop_affected_wui.csv')),
  'non-WUI' = read_csv(here('data/processed/pop_affected_non_wui.csv')),
  .id = 'pop_wui'
) %>%
  select(-ID_hazard) %>%
  rename(wildfire_year = year)
pop_affected_per_fire <- bind_rows(
  'WUI' = read_csv(here('data/processed/pop_affected_per_fire_non_wui.csv')),
  'non-WUI' = read_csv(here('data/processed/pop_affected_per_fire_wui.csv')),
  .id = 'pop_wui'
) %>%
  mutate(ID_hazard = as.integer(ID_hazard)) %>%
  rename(wildfire_id = ID_hazard, wildfire_year = year)
```

Important to note here that we don't acually know where fatalities happened or structures were
destoryed, so we can only categorize those by whether the wildfire burn zone itself occurred in
the WUI.

Population affected, however, is categorized by whether the population itself was in the WUI
regarless of the fire's designation.

## Structures destroyed and fatalities

Summarized according to whether fire was in WUI.

```{r table}
wfbz_by_region <- wfbz %>%
  st_drop_geometry() %>%
  mutate(
    usfs_region = wfbz_region,
    wildfire_wui = if_else(is.na(wildfire_wui), 'non-WUI', 'WUI')
  ) %>%
  filter(
    wildfire_community_intersect
  ) %>%
  bind_rows(., mutate(., usfs_region = 'Nationwide'))

summ_by_region <- wfbz_by_region %>%
  group_by(usfs_region, wildfire_wui) %>%
  summarize(
    across(
      c(wildfire_struct_destroyed, wildfire_civil_fatalities),
      .fns = c(
        `mean` = \(.x) mean(.x, na.rm = TRUE),
        `SD` = \(.x) sd(.x, na.rm = TRUE),
        `25th pctl.` = \(.x) quantile(.x, .25, na.rm = TRUE),
        `median` = \(.x) quantile(.x, .5, na.rm = TRUE),
        `75th pctl.` = \(.x) quantile(.x, .75, na.rm = TRUE)
      ),
      .names = '{.fn} {.col}'
    )
  ) %>%
  left_join(REGION_ORDER) %>%
  arrange(usfs_region != 'Nationwide', desc(plot_order)) %>%
  select(-plot_order)

gt(summ_by_region)
```


## Population Affected

All community-intersecting fires regardless of WFBZ's WUI designation.

```{r}
summ_by_region <- wfbz_by_region %>%
  left_join(pop_affected_per_fire) %>%
  group_by(usfs_region, pop_wui) %>%
  summarize(
    across(
      c(`Pop. exposed` = exposed_10km),
      .fns = c(
        `mean` = \(.x) mean(.x, na.rm = TRUE),
        `SD` = \(.x) sd(.x, na.rm = TRUE),
        `25th pctl.` = \(.x) quantile(.x, .25, na.rm = TRUE),
        `median` = \(.x) quantile(.x, .5, na.rm = TRUE),
        `75th pctl.` = \(.x) quantile(.x, .75, na.rm = TRUE)
      ),
      .names = '{.fn} {.col}'
    )
  ) %>%
  left_join(REGION_ORDER) %>%
  arrange(usfs_region != 'Nationwide', desc(plot_order)) %>%
  select(-plot_order)

gt(summ_by_region)
```
